---
title: "FAME dynamic"
author: "Martí Puchal Batriu"
date: "2025-12-09"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#library(shinyFiles)
library(dplyr)
library(tidyr)
library(stringr)
library(formattable)

values <- reactiveValues(model_list = c())
tables <- reactiveValues(model_df = data.frame())
data_path <- "../data/"

```


```{r setup functions, include=FALSE}

# Open the different files without shuting the shiny
error_manegement <- function(file_name_opener){
  
  tryCatch({
  # Try block
  # This will cause an error
    x <- read.csv(file_name_opener)

  return(x)
}, error = function(e) {
  # Except block
  message("Error occurred: ", e$message)
  return(Inf)  # Default value
}, warning = function(w) {
  # Warning handler
  message("Warning: ", w$message)
}, finally = {
  # Finally block (always runs)
  message("File Opened")
})
  return("The file was not correcly oppened")
}


# Process the raw df from the model function
df_gen <- function(df_input){
  df_input <- df_input %>% select(ends_with("RT"))
  met_202X <- gsub(".Results_RT", "",colnames(df_input))
  
  vec <- lapply(colnames(df_input),function(x){
  s <- str_split(x, "[^0-9]+", simplify = TRUE)
  return(s[2:3])})
  
  new_cols <- as.data.frame(do.call(rbind, vec))
  colnames(new_cols) <-  c("Carbon","Insat")
  new_cols$RT <- t(df_input)[,1]
  df_output <- new_cols
  row.names(df_output) <- met_202X
  df_output[] <- lapply(df_output, as.numeric)
  return(df_output)
}

# Function to run the different models
values$RT_pred  <- function(Carbons, Insaturation, model_id){
  if (Insaturation==0){
    sample_rt <- as.numeric(predict(values$SFA_lm[[model_id]], data.frame("Carbon"=c(Carbons))))
    
  } else if(Insaturation==1){
    sample_rt <- as.numeric(predict(values$MUFA_lm[[model_id]], data.frame("Carbon"=c(Carbons))))
  } else{
    if (Carbons == 18){
      sample_rt <- as.numeric(predict(values$C18_lm[[model_id]], data.frame("Insat"=c(Insaturation))))
    } else if (Carbons == 20){
      sample_rt <- as.numeric(predict(values$C20_lm[[model_id]], data.frame("Insat"=c(Insaturation))))
    } else if (Carbons == 22){
      sample_rt <- as.numeric(predict(values$C22_lm[[model_id]], data.frame("Insat"=c(Insaturation))))
    } else{
      sample_rt <- "na"
    }
  }
  return(sample_rt)
}


```



# Input dataframe 

A dataframe is used to buid the custom models. This dataframe contains the Fatty Acid with their retaintions times. The input of the dataframe must have the following format: FA Retention Times labeled as FAXX:X_RT in the column names and the RT in the row. And as a difference between the model dataframe and the sample dataframe is the model datafraem have only one row.

File path example: ../data/250312_FAME2023_int_FAMEMIX_RT.csv


```{r open models, echo=FALSE}

fluidPage(
  fluidRow(
    column(12,textAreaInput("model_load",label = "DataFrame to build the model",
                            placeholder = "Enter the path to the file....",
                            width = "100%",
                            height = "35px",
                            resize = "vertical")
           
    )
  ),
  fluidRow(
    column(12,actionButton("model_open",label = "Open")
           
    )
  ),
  tags$br(),
  tags$br()
)




observeEvent(input$model_open,{
  df <- error_manegement(input$model_load)
  if (is.data.frame(df)){
              tables$model_df <- df_gen(df)
}else{tables$model_df <- df}
              })



renderTable(tables$model_df,rownames = TRUE)


```

```{r build model list, echo=FALSE}

fluidPage(
  fluidRow(
    column(12,
      textAreaInput("model_name", 
                   label = "Name the model", 
                   placeholder = "Enter model name here...",
                   width = "50%",
                   height = "50px",
                   resize = "vertical")
    )
  ),
  fluidRow(
    column(12,
      style = "margin-top: 15px;",
      actionButton("add_model", label = "Add", 
                  style = "margin-right: 10px;"),
      actionButton("Clear_model", label = "Clear",
                  style = "margin-right: 10px;"),
      actionButton("Load_defauts", label = "Load default models")
    )
  ),
  tags$br(),
  tags$br()
)



observeEvent(input$Clear_model, 
              {
                values$model_list <- c()
                print(getwd())
                })

observeEvent(input$add_model,{
  
  if (is.data.frame(tables$model_df)){if (ncol(tables$model_df)>=1){
  data_set <- tables$model_df
  
  SFA <- data_set[data_set$Insat==0,]
  values$SFA_lm[[input$model_name]] <- lm(RT~Carbon,SFA)
  
  MUFA <- data_set[data_set$Insat==1,]
  values$MUFA_lm[[input$model_name]] <- lm(RT~Carbon,MUFA)
  
  C18 <- data_set[data_set$Carbon==18,]
  values$C18_lm[[input$model_name]] <- lm(RT~Insat,C18)
  
  C20 <- data_set[data_set$Carbon==20,]
  values$C20_lm[[input$model_name]] <- lm(RT~Insat,C18)
  
  C22 <- data_set[data_set$Carbon==22,]
  values$C22_lm[[input$model_name]] <- lm(RT~Insat,C18)
  
  values$model_list <- append(values$model_list,input$model_name)
  }}
                           
  })
observeEvent(input$Load_defauts,{
  
  RT_FA <- read.delim(paste0(data_path,"RT_FA.tsv"))
  RT_FA_2023 <- read.csv(paste0(data_path,"250312_FAME2023_int_FAMEMIX_RT.csv"))
  RT_FA_2024 <- read.csv(paste0(data_path,"250312_FAME2024_int_FAMEMIX_RT.csv"))

  # Process df function
  df_gen <- function(df_input){
  df_input <- df_input %>% select(ends_with("RT"))
  met_202X <- gsub(".Results_RT", "",colnames(df_input))
  
  vec <- lapply(colnames(df_input),function(x){
  s <- str_split(x, "[^0-9]+", simplify = TRUE)
  return(s[2:3])})
  
  new_cols <- as.data.frame(do.call(rbind, vec))
  colnames(new_cols) <-  c("Carbon","Insat")
  new_cols$RT <- t(df_input)[,1]
  df_output <- new_cols
  row.names(df_output) <- met_202X
  df_output[] <- lapply(df_output, as.numeric)
  return(df_output)
  }

  RT_FA_2023 <- df_gen(RT_FA_2023)
  RT_FA_2024 <- df_gen(RT_FA_2024)
  
  count = 0
  data_names <- c("RT_FA","RT_FA_2023","RT_FA_2024")

for (data_set in list(RT_FA,RT_FA_2023,RT_FA_2024)){
  
  # Build the linear regresion for each of the datasets 
  count = count+1
  SFA <- data_set[data_set$Insat==0,]
  values$SFA_lm[[data_names[count]]] <- lm(RT~Carbon,SFA)
  
  MUFA <- data_set[data_set$Insat==1,]
  values$MUFA_lm[[data_names[count]]] <- lm(RT~Carbon,MUFA)
  
  C18 <- data_set[data_set$Carbon==18,]
  values$C18_lm[[data_names[count]]] <- lm(RT~Insat,C18)
  
  C20 <- data_set[data_set$Carbon==20,]
  values$C20_lm[[data_names[count]]] <- lm(RT~Insat,C18)
  
  C22 <- data_set[data_set$Carbon==22,]
  values$C22_lm[[data_names[count]]] <- lm(RT~Insat,C18)
  
  values$model_list <- append(values$model_list,data_names[count])
}
  
})

reactive(
if (length(values$model_list)==0 & is.null(values$model_list)){
    values$n=1
    
  }else{
    values$n=1:length(values$model_list)
    
  }
)

fluidPage(
  
  
renderPrint(
            paste0(
              paste0(
                  paste0(
                    "Model ",
                    values$n
                    ),": "
                  ),
                  values$model_list
              ))
)



```

# Load sample data


```{r sample opening,echo=FALSE}
fluidPage(
  fluidRow(
    column(12,textAreaInput("sample_load",label = "DataFrame to analyse",
                            placeholder = "Enter the path to the file....",
                            width = "100%",
                            height = "35px",
                            resize = "vertical"))
  ),
  fluidRow(
    column(12,actionButton("sample_open",label = "Open"))
  )
)

observeEvent(input$sample_open,{
              tables$sample_df <- error_manegement(input$sample_load)})

renderTable(tables$sample_df,rownames = TRUE)
```


# Analyse the data

```{r,echo=FALSE}
fluidPage(
  fluidRow(
    column(12,selectInput("choices_models", label = "Choose the model to work with",
              choices = NULL)
    )
    
  ),
  fluidRow(
    column(12,actionButton("analyse_sample",label = "Analyse"))
  )
)




model_names <- reactive({
    values$model_list})

observeEvent(model_names(), {
    updateSelectInput(session, 
                      "choices_models",
                      choices = model_names())
  })


observeEvent(input$analyse_sample,{
  
  if (!is.null(tables$sample_df) & is.data.frame(tables$sample_df)){

area_df <- tables$sample_df %>% select(ends_with("Area"))

area_sum <- apply(area_df,1,sum)
FA16.0_prop <- tables$sample_df[grep("FA16\\.0.+Area",colnames(tables$sample_df),value = TRUE)]/area_sum
FA16.0_prop[FA16.0_prop>=0.2] <- 0.2
FA16.0_prop <- as.numeric(FA16.0_prop[,1])
color_palette <- colorRampPalette(c( "red","white"))(20)


format_list <- list(
  FA16.0.Results_Area	 = formatter("span",
    style = x ~ {
      style(
        display = "block",
        "border-radius" = "4px",
        "background-color" = color_palette[as.integer(FA16.0_prop * 100)]
      )
    }
  )
)    

rt_fa_names <- grep("RT",colnames(tables$sample_df),value = TRUE)

CI <- function(rt_names){
  vec <- lapply(rt_names,function(x){
    s <- str_split(x, "[^0-9]+", simplify = TRUE)
    return(s[2:3])})

  new_cols <- as.data.frame(do.call(rbind, vec))
  colnames(new_cols) <-  c("Carbon","Insat")
  row.names(new_cols) <- rt_names
  
  return(new_cols)
}


ratios_rt <- sweep(tables$sample_df[rt_fa_names], 2, apply(CI(rt_fa_names),1,function(x){values$RT_pred(as.numeric(x[1]),as.numeric(x[2]),input$choices_models)}), "/")

color_list <- list()
count <- 0

for (cols in colnames(ratios_rt)){
  count = count + 1
  
  color_list[[count]] <- ifelse(ratios_rt[[cols]] > 1.5, "red",
                                ifelse(ratios_rt[[cols]] < 0.5,"blue","black"))
  
  format_list[[cols]] <- local({
    current_count <- count  # Capture current value
    formatter("span", style = x ~ style("color" = color_list[[current_count]]))
  })
}

output$formatted_table <- renderFormattable({
    formattable(tables$sample_df, format_list)
  })
}
})

formattableOutput("formatted_table")

```











